---
title: "Preliminary exploration"
author: "Claire Levy"
date: "Monday, March 09, 2015"
output:
beamer_presentation:
slide_level:2
---

```{r, echo=FALSE, message=FALSE}

setwd("J:/MacLabUsers/Claire/Projects/HVE-microarray/differentiallyExpressedGenes")
require(dplyr)
require(reshape2)
require(stringr)
require(ggplot2)
require(pander)
dataSummary<-read.csv("ex_vivo_HVE_Tenofovir_Summary_results.csv")

#spread sheet is laid out with the 4 columns: log2 fold change
#raw fold change, FDR, Differentially expressed gene (DEG)
#These 4 columns are repeated for the 4 time points and for the
#two doses of tenofovir (500 and 50)

#timepoints are D1, D4, D7, D14



#I am going to isolate just the DEG summary data from the end
#of the spreadsheet (columns 36-43) and the annotations in the first
#three columns and then
#extract the probe, target and Entrez IDs for the
#DEG for each time point in each direction (up or down)


summaryOnly<-dataSummary[,c(1:3,36:43)]

names(summaryOnly)[4:11]<-c("d14.50","d7.50","d4.50","d1.50",
                            "d14.500","d7.500","d4.500","d1.500")



##melt and show HOW MANY are up, down and false for each sample
melted<-melt(summaryOnly,
             id.vars=c("Probe.ID","TargetID","ENTREZ_GENE_ID"),
             variable.name="day.dose",value.name="UpDownFalse")

#change order of samples for future plots
melted$day.dose<-factor(melted$day.dose,
                        levels=c("d1.50","d4.50","d7.50",
                                 "d14.50","d1.500","d4.500","d7.500",
                                 "d14.500"))

upDownCount<-melted%>%
  group_by(day.dose)%>%
  summarise(countUP=sum(UpDownFalse=="UP"),
            countDOWN=sum(UpDownFalse=="DOWN"),
            countFALSE=sum(UpDownFalse=="FALSE"))
```


##I've done a few things with the data so far, focusing mostly on upregulated genes for dose = 50:

### Read in DEG data from excel, rearrange a little and calculate  number of up, down, and non-regulated genes per day for dose = 50.

### Extract data for each day for dose = 50 for upregulated genes.

### Run those data through DAVID and find GO terms upregulated genes that were common to all days at dose = 50.
----

```{r, echo=FALSE, message=FALSE}
require(pander)
pander(upDownCount)
```
----

```{r, echo=FALSE}
#now melt that df to prepare it for easy plotting
meltedupDownCount<-melt(upDownCount, id.vars="day.dose",
                        variable.name="Direction",value.name="Count")


#how many genes are there total?
nGenes<-length(summaryOnly$Probe.ID)

```
----
```{r, echo=FALSE}
##plot of up down and false counts
ggplot(meltedupDownCount,aes(x=day.dose,y=Count))+
  geom_bar(aes(fill=Direction), position="dodge",stat="identity")+ylab("Number of genes")+
  geom_hline(y=7832)+
  ggtitle("Number of upregulated, downregulated and non-significant genes\n\ per time point and dose\n\ 
          (line represents total genes analyzed for diff exp)")
```
----
##Sean and I wrote some functions to arrange the data, extract days and doses of interest in formats that DAVID and InnateDB can use, and then read in the saved data from DAVID.

```{r, echo=FALSE}


#using the function to make data in longform...
toLongForm<-source("DEG-to-long-form.R")

#a function to extract and write data from specific days and concentrations
#the table is for innateDB.
#the csv is for david where you just put in the IDs
#For innateDB upload: first col is entrez, second is log2foldchange, 3rd is adj pval)

extract<-function(day,concentration,direction){
x<-longForm[longForm$Day == day &
              
              longForm$Concentration == concentration &
              
              longForm$Direction == direction,
            c(3,6,8)]

write.table(x,file=paste(day,concentration,direction,"txt",sep="."),
         sep="\t",row.names=FALSE,col.names=FALSE)

write.csv(x[,1],file=paste(day,concentration,direction,"csv",sep="."),
          row.names=FALSE)
}



setwd("J:/MacLabUsers/Claire/Projects/HVE-microarray/differentiallyExpressedGenes")
#all the UPs from concentration=50
extract("1","50","UP")
extract("4","50","UP")
extract("7","50","UP")
extract("14","50","UP")


#get DAVID data (saved as tab delim txt)
#left the defaults but chose "Fisher's exact" under options
# and chose to include a column for Foldchange

#here is function for getting the DAVID data out of the folder
#I also added a column to each df saying which day it represented
getDAVID<-function(day, concentration, direction){
 y<-read.table(file=paste("DAVID",day, concentration,direction,"txt", sep="."),
               header=TRUE,sep="\t")
}



DAVID.1.50.UP<-getDAVID("1","50","UP")
DAVID.1.50.UP<-DAVID.1.50.UP%>%
  mutate(Day= rep("1", times=nrow(DAVID.1.50.UP)))


DAVID.7.50.UP<-getDAVID("7","50","UP")
DAVID.7.50.UP<-DAVID.7.50.UP%>%
  mutate( Day= rep("7", times=nrow(DAVID.7.50.UP)))

DAVID.14.50.UP<-getDAVID("14","50","UP")
DAVID.14.50.UP<-DAVID.14.50.UP%>%
  mutate(Day = rep("14", times=nrow(DAVID.14.50.UP)))

#combine all the data frames EXCEPT day 4 because of weirdness
allDAVID.50.UP<-rbind(DAVID.1.50.UP,DAVID.7.50.UP,DAVID.14.50.UP)
allDAVID.50.UP$Day<-factor(allDAVID.50.UP$Day, levels=c("1","7","14"))
#make a vector of the terms in Day1 UP
DAVID.1.50.UPterms<-DAVID.1.50.UP%>%
  select(Term)

#subset for just the terms that are in Day 1
#notice that I need to specify $Term in DAVID.1.50.UPterms
#even those there is just the one column

toKeep<-allDAVID.50.UP$Term %in% DAVID.1.50.UPterms$Term
overlapTerms.50.UP<-allDAVID.50.UP[toKeep,]

#let's just look at the terms,the pvals,count and fold enrich
#just the ones that are there on all days.


####DF showing overlapping terms for dose 50 days 1,7,14###
##note that the associated gene IDS are not in this df but they are
#in allDAVID.50.UP
overlapTerms.50.UPShort<-overlapTerms.50.UP%>%
  select(Day,Category,Term,PValue,Count, Fold.Enrichment)%>%
  group_by(Term)%>%
  filter(n()==3)%>%
  arrange(Day,Term,PValue)
```
----
##DAVID terms that overlap in days 1,7,14 for upregulated genes, dose = 50.

```{r, echo=F}
pander(overlapTerms.50.UPShort)
#if you do summarize(n())where filter is  it shows you the term and 
#how many occurances there were in overlapTerms.50.UP. We only want the terms where
#there there 3 occurances (1 per day we looked at)

```
---
##Plots of pvalues, fold enrichment and gene count for the same data set as above.

```{r, echo=FALSE}
###plot plot plot
#note that I changed the x axis labels to a shortened version
# of the term for readability

#Pvalues
ggplot(data=overlapTerms.50.UPShort, aes())+
  geom_point(aes(x = Term , y =PValue,color=Day),
             position=position_jitter(w=0.15),size=4)+
  scale_x_discrete(labels=c("Acetylation","Cytoskeleton",
                            "Nuclear body","Nuclear speck"))+
  ggtitle("GSEA P-values for overlapping GO terms \n\ in Up-regulated genes,dose=50")
  
```
---
##Fold enrichment
```{r, echo=FALSE}
#Fold enrichment
ggplot(data=overlapTerms.50.UPShort, aes())+
  geom_point(aes(x = Term , y =Fold.Enrichment,color=Day),
             position=position_jitter(w=0.15),size=4)+
  scale_x_discrete(labels=c("Acetylation","Cytoskeleton",
                            "Nuclear body","Nuclear speck"))+
  ggtitle("Fold Enrichment for overlapping GO terms \n\ in Up-regulated genes,dose=50")
```
---
#Gene Count
```{r,echo=FALSE}
#Count of genes relating to the term

ggplot(data=overlapTerms.50.UPShort, aes())+
  geom_point(aes(x = Term , y =Count,color=Day),
             position=position_jitter(w=0.15),size=4)+
  scale_x_discrete(labels=c("Acetylation","Cytoskeleton",
                            "Nuclear body","Nuclear speck"))+
  ggtitle("Number of up-regulated genes associated with overlapping terms \n\
dose=50")
```
